Using the UK Biobank as a global reference of worldwide populations: application to measuring ancestry diversity from GWAS summary statistics

### 2021-10-27 paper-ancestry-prop.tex
reexport code

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=\textwidth}\tabbeg}{\tabend\end{adjustbox}}

@@

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Fri Oct 22 15:40:44 2021
\begin{table}[!htb]
	\centering
	\begin{tabular}{|l|c||c|c|c|c|c||c|c|c|c|c||c|c|c|c|c|c|c|c|}
		\hline
		Population & N & BBJ & FinnGen & Per\'u & Qatar & Africa & UKBB & GERA & PAGE & AMR & Caribbean & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
		\hline
		Africa (West) & 735 &  &  &  &  & 34.4 & 1.2 & 1.7 & 22.6 & 1.3 & 72.5 & 0.2 & 0.2 & 2 & 0.6 & 3.4 & 0.3 & 0.5 & 2.2 \\ 
		Africa (Central) & 276 &  &  &  &  & 4.4 & 0.3 & 0.5 & 6.9 & 0.1 & 15.1 & 0.1 &  & 0.7 & 0.7 & 1.1 & 0.1 & 0.2 & 2.3 \\ 
		Africa (South) & 449 &  &  &  &  & 61.1 & 0.2 & 0.2 & 3 & 0.5 & 3.6 &  & 0.1 & 0.4 & 0.8 & 0.5 & 0.1 &  & 2.3 \\ 
		Africa (East 2) & 279 &  &  &  &  &  &  &  & 0.7 &  & 1 &  &  & 0.3 & 0.1 & 0.3 &  &  & 1.1 \\ 
		\hline
		Africa (East 1) & 276 &  &  &  & 11.4 &  &  &  &  &  &  &  &  &  & 0.3 & 0.1 &  &  & 1.1 \\ 
		Africa (North) & 268 &  &  &  & 22.1 &  &  &  & 1.5 & 1.1 &  &  &  &  &  &  &  &  &  \\ 
		Middle East & 523 &  &  &  & 63.5 &  &  & 0.9 &  &  &  &  &  & 2.8 & 0.1 & 0.1 &  & 0.7 & 1 \\ 
		\hline
		Ashkenazi & 1975 &  &  &  & 1.9 &  & 0.9 & 5 & 0.4 &  & 0.1 & 0.3 & 2.3 & 1.3 & 1.3 & 0.6 & 0.8 & 2.3 & 0.8 \\ 
		Italy & 345 &  &  &  &  &  & 2.7 & 6 & 2.3 & 0.9 & 0.1 & 4 & 4.3 & 3.4 & 4.5 & 1 & 4.1 & 6.4 & 3.1 \\ 
		Europe (East) & 1223 &  & 7.5 &  &  &  & 3 & 9.8 & 0.5 &  &  & 11.9 & 12.7 & 13.2 & 15 & 15 & 12.9 & 11.8 & 10.2 \\ 
		Finland & 242 &  & 85.8 &  &  &  &  & 1.3 & 0.5 & 0.2 &  & 9 & 12.2 & 6.1 & 8 & 13 & 10.4 & 4.7 & 2.4 \\ 
		Europe (North West) & 4416 &  & 6 &  &  &  & 81.2 & 53.8 & 8.4 &  & 6.2 & 62.8 & 58.4 & 47.5 & 50.6 & 56.5 & 63.5 & 62.3 & 37.8 \\ 
		Europe (South West) & 603 &  &  &  &  &  & 7.7 & 8.4 & 8.2 & 6.7 & 1 & 7 & 7.1 & 4.3 & 5 & 3.1 & 6.4 & 7.5 & 4.3 \\ 
		\hline
		South America & 473 &  & 0.3 & 94.7 &  &  & 0.2 & 5.1 & 27.7 & 89.0 & 0.3 & 1.1 & 0.8 & 1.6 & 0.8 & 2.1 & 0.6 & 0.3 & 0.3 \\ 
		\hline
		Bangladesh & 309 &  &  &  &  &  & 0.6 & 0.3 & 1.3 &  &  & 1.2 & 1 & 2.7 & 2.7 & 0.9 &  &  & 1.3 \\ 
		Pakistan & 400 &  &  &  &  &  & 1.4 &  & 0.1 &  &  & 0.6 &  & 6.6 & 3.1 & 1.1 &  &  & 0.3 \\ 
		Sri Lanka & 372 &  &  &  & 1.1 &  & 0.3 & 0.1 & 0.7 &  &  & 0.9 & 0.4 & 2.2 & 1.6 & 0.3 &  &  & 0.7 \\ 
		\hline
		Philippines & 295 &  &  &  &  &  & 0.1 & 1.4 & 3.9 &  &  &  &  &  &  & 0.2 & 0.1 & 0.1 & 0.5 \\ 
		Asia (East) & 961 & 3.9 &  &  &  &  & 0.3 & 3.8 & 2.6 &  &  & 0.6 & 0.3 & 2.5 & 1.8 & 0.3 &  & 3.1 & 2.4 \\ 
		Japan & 344 & 96.1 & 0.5 & 5.3 &  &  &  & 1.9 & 8.7 &  &  & 0.3 & 0.3 & 2.2 & 3.1 & 0.5 & 0.6 &  & 26 \\ 
		\hline
	\end{tabular}
	\caption{Reference population with their size (N) and corresponding ancestry proportions in several GWAS datasets. Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), the UK Biobank (UKBB, \cite{bycroft2018uk}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}. Several of them have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}. \label{tab:table1}}
\end{table}

@@

We download several GWAS summary statistics which reports allele frequencies and apply this method to both validate our new reference set we provide and show one of its potential use.
We first apply function \texttt{snp\_ancestry\_summary} to more homogeneous samples as an empirically validation.
When applying this function (using the new reference set we provide) to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions estimates obtained match expectations (Table \ref{tab:table1}).
When comparing our estimates with reported ancestries, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries (self-reported), whereas our estimates are of 27.7\% South American, 20.2\% European, 34.8\% African, 2.1\% South Asian, 11.3\% East Asian, and 3.9\% Filipino [TODO: CHANGE].
GWAS summary statistics from either European ancestries or trans-ethnic ancestries all have a substantial proportion estimated from European ancestry groups, while ancestries from other continents are still largely underrepresented (Table \ref{tab:table1}).

Here we provide an unprecedentedly large and diverse reference set of allele frequencies using the UK Biobank.
We then show how to effectively measure diversity from GWAS summary statistics using this new set of reference allele frequencies.
Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
Our work has limitations though. 
First, although we managed to define ancestry groups with more than 240 individuals, it seems that these groups are still not large enough for the ancestry deconvolution method to recover 100\% of one ancestry (e.g.\ the case for the FinnGen and Biobank Japan cohorts which are likely very homogenous). In this case, we also capture a small proportion from nearby ancestry groups. 
Second, it is not fully clear whether we can capture any ancestry, e.g.\ Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
Third, with the 17 ancestry groups we defined, we think we are capturing a large proportion of the genetic diversity in both Europe and Africa, but we may still be missing more fine-grained diversity in both South Asia and South America, as well as in some isolated populations.
Fourth, when we use the frequencies reported in the GWAS summary statistics, it is not clear whether they were recomputed from all individuals (e.g.\ before performing any quality control and filtering), and for meta-analyses of binary traits, whether they were computed as weighted average using total sample sizes or effective sample sizes.
Anyway, despite these limitations, we envision that the new set of reference allele frequencies we provide will have many useful applications.

@@

The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
The set of reference allele frequencies for 5,816,590 genetic variants across 17 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31163785} [TODO: REUPLOAD AND VERIFY].
All code used for this paper is available at \url{https://github.com/privefl/freq-ancestry/tree/main/code}. 
We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.

>>>>

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}

@@

@@

We download several GWAS summary statistics which reports allele frequencies and apply this method to both validate our new reference set we provide and show one of its potential use.
We first apply function \texttt{snp\_ancestry\_summary} to more homogeneous samples as an empirically validation.
When applying this function (using the new reference set we provide) to the Biobank Japan (Japanese cohort), FinnGen (Finnish), a Peruvian cohort, a Qatari cohort and Sub-Saharan African cohort, the ancestry proportions estimates obtained match expectations (Table \ref{tab:table1}).
When comparing our estimates with reported ancestries, for example PAGE is composed of 44.6\% Hispanic-Latinos, 34.7\% African-Americans, 9.4\% Asians, 7.9\% Native Hawaiians and 3.4\% of some other ancestries (self-reported), whereas our estimates are of 24.9\% South American, 21.9\% European, 35.4\% African, 2.0\% South Asian, 11.3\% East Asian, 3.9\% Filipino, and 0.5\% Middle East.
GWAS summary statistics from either European ancestries or trans-ethnic ancestries all have a substantial proportion estimated from European ancestry groups, while ancestries from other continents are still largely underrepresented (Table \ref{tab:table1}).

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Wed Oct 27 13:40:20 2021
\begin{table}[htb]
	\centering
	\begin{tabular}{|l|c||c|c|c|c|c||c|c|c||c|c|c|c|c|c|c|c|}
		\hline
		Population & N & BBJ & FinnGen & Per\'u & Qatar & Africa & UKBB & GERA & PAGE & BrCa & PrCa & CAD & body fat & covid & eczema & epilepsy & urate \\ 
		\hline
		Africa (West) & 735 &  &  &  &  & 36.7 & 1.4 & 1.9 & 26.4 & 0.2 & 0.2 & 2.4 & 1 & 4 & 0.3 & 0.6 & 3.6 \\ 
		Africa (South) & 449 &  &  &  &  & 63.3 & 0.3 & 0.4 & 6.8 &  & 0.1 & 0.9 & 1.1 & 1.2 & 0.1 &  & 4 \\ 
		Africa (East) & 276 &  &  &  & 11.4 &  &  &  & 0.2 &  &  &  & 0.2 & 0.1 &  &  & 1.4 \\ 
		Africa (North) & 268 &  &  &  & 22.1 &  &  & 0.4 & 2 &  &  &  &  &  &  & 0.3 &  \\ 
		\hline
		Middle East & 523 &  &  &  & 63.5 &  & 1.3 & 3.4 & 0.5 & 2 & 2.2 & 5.2 & 2.9 & 1.5 & 1.6 & 3.2 & 2.8 \\ 
		\hline
		Ashkenazi & 1975 &  &  &  & 1.9 &  & 1.4 & 6 & 0.9 & 1.3 & 3.3 & 2.2 & 2.3 & 1.3 & 2 & 3.6 & 1.5 \\ 
		Europe (North East) & 320 &  & 8.9 &  &  &  & 3.8 & 7.7 & 0.7 & 9.6 & 10 & 10 & 11 & 11.7 & 10.1 & 8.9 & 7.5 \\ 
		Finland & 143 (+ 99) &  & 85.2 &  &  &  & 2.7 & 3.5 & 0.7 & 11.9 & 14.8 & 8.2 & 10.5 & 15.4 & 13.1 & 7.1 & 4.3 \\ 
		British-Irish Isles & 4000 &  & 5.1 &  &  &  & 79.1 & 53 & 8.9 & 59.8 & 56.1 & 46.6 & 49.2 & 54.8 & 61.7 & 62 & 36.6 \\ 
		Europe (South West) & 603 &  &  &  &  &  & 7.7 & 12.2 & 10.7 & 11.2 & 11.2 & 7.5 & 9.5 & 5.3 & 10.1 & 11.2 & 7.3 \\ 
		\hline
		South America & 473 (+ 84) &  & 0.4 & 99.1 &  &  & 0.1 & 4.5 & 24.9 & 0.9 & 0.6 & 1.4 & 0.7 & 1.8 & 0.4 & 0.1 & 0.2 \\ 
		\hline
		Japan & 240 (+ 104) & 96.1 & 0.4 & 0.9 &  &  &  & 1.8 & 8.8 & 0.2 & 0.2 & 2.1 & 3 & 0.4 & 0.4 &  & 25.9 \\ 
		Asia (East) & 961 & 3.9 &  &  &  &  & 0.2 & 3.7 & 2.5 & 0.6 & 0.3 & 2.5 & 1.8 & 0.2 &  & 3 & 2.3 \\ 
		Philippines & 295 &  &  &  &  &  & 0.1 & 1.4 & 3.9 &  &  &  &  & 0.2 & 0.1 & 0.1 & 0.5 \\ 
		\hline
		Bangladesh & 223 (+ 86) &  &  &  &  &  & 0.6 &  & 1.2 & 1.1 & 0.8 & 2.8 & 2.8 & 0.9 &  &  & 1.4 \\ 
		Sri Lanka & 372 &  &  &  & 1.1 &  & 0.4 &  & 0.8 & 0.8 & 0.2 & 2.2 & 1.6 & 0.4 &  &  & 0.7 \\ 
		Pakistan & 400 &  &  &  &  &  & 0.8 &  &  & 0.4 &  & 6 & 2.5 & 0.7 &  &  &  \\ 
		\hline
	\end{tabular}
	\caption{Reference population with their size (N) and corresponding ancestry proportions in several GWAS datasets. Citations for the allele frequencies used: the Biobank Japan (BBJ, \cite{sakaue2021cross}), FinnGen (\url{https://r5.finngen.fi/}), GWAS in Peruvians \cite[]{asgari2020positively}, GWAS in Qataris \cite[]{thareja2021whole}, GWAS in Sub-Saharan Africans (Africa, \cite{chen2019genome}), the UK Biobank (UKBB, \cite{bycroft2018uk}), GERA \cite[]{hoffmann2018large}, PAGE \cite[]{wojcik2019genetic}, breast cancer (BrCa, \cite{michailidou2017association}), prostate cancer (PrCa, \cite{schumacher2018association}), coronary artery disease (CAD, \cite{nikpay2015comprehensive}), body fat percentage \cite[]{lu2016new}, COVID-19 \cite[]{ganna2021mapping}, eczema \cite[]{paternoster2015multi}, epilepsy \cite[]{consortium2018genome}, and serum urate \cite[]{tin2019target}. Several of them have been downloaded through the NHGRI-EBI GWAS Catalog \cite[]{macarthur2017new}. \label{tab:table1}}
\end{table}

Here we provide an unprecedentedly large and diverse reference set of allele frequencies using the UK Biobank.
We then show how to effectively measure diversity from GWAS summary statistics using this new set of reference allele frequencies.
Measuring genetic diversity is an important problem because increasing genetic diversity is key to making new genetic discoveries, while also being a major source of confounding to be aware of in genetics studies.
Our work has limitations though. 
First, although we managed to define ancestry groups with more than 240 individuals, it seems that these groups are still not large enough for the ancestry deconvolution method to recover 100\% of one ancestry (e.g.\ the case for the FinnGen and Biobank Japan cohorts which are likely very homogenous). In this case, we also capture a small proportion from nearby ancestry groups. 
Second, it is not fully clear whether we can capture any ancestry, e.g.\ Native Hawaiians in the PAGE study; it seems that they are partly captured by the ``Philippines'' ancestry group we define.
Third, with the 17 ancestry groups we defined, we think we are capturing a large proportion of the genetic diversity in both Europe and Africa, but we may still be missing more fine-grained diversity in both South Asia and South America.
Fourth, when we use the frequencies reported in the GWAS summary statistics, it is not clear whether they were recomputed from all individuals (e.g.\ before performing any quality control and filtering), and for meta-analyses of binary traits, whether they were computed as weighted average using total sample sizes or effective sample sizes.
Anyway, despite these limitations, we envision that the new set of reference allele frequencies we provide will have many useful applications.

@@

The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
The set of reference allele frequencies for 5,816,590 genetic variants across 17 diverse ancestry groups we create here can be downloaded from \url{https://figshare.com/ndownloader/files/31218445}.
All code used for this paper is available at \url{https://github.com/privefl/freq-ancestry/tree/main/code}. 
We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
